---
title: "Maximum Likelihood Theory"
draft: false
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: STAT 4690--Applied Multivariate Analysis
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
  - \usepackage{xcolor}
---

```{r,setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Sufficient Statistics {.allowframebreaks}

  - We saw in the previous lecture that the multivariate normal distribution is completely determined by its mean vector $\mu \in\mathbb{R}^p$ and its covariance matrix $\Sigma$.
  - Therefore, given a sample $\mathbf{Y}_1, \ldots, \mathbf{Y}_n\sim N_p(\mu, \Sigma)$ ($n > p$), we only need to estimate $(\mu,\Sigma)$. 
    + Obvious candidates: sample mean $\bar{\mathbf{Y}}$ and sample covariance $S_n$.
  - Write down the *likelihood*:
  \begin{align*} 
  L &= \prod_{i=1}^n\left(\frac{1}{\sqrt{(2\pi)^p\lvert\Sigma\rvert}}\exp\left(-\frac{1}{2}(\mathbf{y}_i - \mu)^T\Sigma^{-1}(\mathbf{y}_i - \mu)\right)\right)\\
   &= \frac{1}{(2\pi)^{np/2}\lvert\Sigma\rvert^{n/2}}\exp\left(-\frac{1}{2}\sum_{i=1}^n(\mathbf{y}_i - \mu)^T\Sigma^{-1}(\mathbf{y}_i - \mu)\right)
  \end{align*}
  - If we take the (natural) logarithm of $L$ and drop any term that does not depend on $(\mu,\Sigma)$, we get
  $$\ell = -\frac{n}{2}\log\lvert\Sigma\rvert - \frac{1}{2}\sum_{i=1}^n(\mathbf{y}_i - \mu)^T\Sigma^{-1}(\mathbf{y}_i - \mu).$$
  - If we can re-express the second summand in terms of $\bar{\mathbf{Y}}$ and $S_n$, by the Fisher-Neyman factorization theorem, we will then know that $(\bar{\mathbf{Y}},S_n)$ is jointly **sufficient** for $(\mu,\Sigma)$.
  - First, we have
  \begin{multline*}
  \sum_{i=1}^n(\mathbf{y}_i - \mu)(\mathbf{y}_i - \mu)^T = \sum_{i=1}^n(\mathbf{y}_i - \bar{\mathbf{y}} + \bar{\mathbf{y}} - \mu)(\mathbf{y}_i - \bar{\mathbf{y}} + \bar{\mathbf{y}} - \mu)^T\\
  = \sum_{i=1}^n\left((\mathbf{y}_i - \bar{\mathbf{y}})(\mathbf{y}_i - \bar{\mathbf{y}})^T + (\mathbf{y}_i - \bar{\mathbf{y}})(\bar{\mathbf{y}} - \mu)^T\right.\\
  \quad \left.+ (\bar{\mathbf{y}} - \mu)(\mathbf{y}_i - \bar{\mathbf{y}})^T + (\bar{\mathbf{y}} - \mu)(\bar{\mathbf{y}} - \mu)^T\right)\\
  = \sum_{i=1}^n(\mathbf{y}_i - \bar{\mathbf{y}})(\mathbf{y}_i - \bar{\mathbf{y}})^T + n(\bar{\mathbf{y}} - \mu)(\bar{\mathbf{y}} - \mu)^T\\
  \quad \quad= (n-1)S_n + n(\bar{\mathbf{y}} - \mu)(\bar{\mathbf{y}} - \mu)^T.\hfill
  \end{multline*}
  - Next, using the fact that $\mathrm{tr}(ABC) = \mathrm{tr}(BCA)$, we have
  \begin{align*}
  \sum_{i=1}^n(\mathbf{y}_i - \mu)^T\Sigma^{-1}(\mathbf{y}_i - \mu) &= \mathrm{tr}\left(\sum_{i=1}^n(\mathbf{y}_i - \mu)^T\Sigma^{-1}(\mathbf{y}_i - \mu)\right) \\
  &= \mathrm{tr}\left(\sum_{i=1}^n\Sigma^{-1}(\mathbf{y}_i - \mu)(\mathbf{y}_i - \mu)^T\right) \\
  &= \mathrm{tr}\left(\Sigma^{-1}\sum_{i=1}^n(\mathbf{y}_i - \mu)(\mathbf{y}_i - \mu)^T\right) \\
  &= (n-1)\mathrm{tr}\left(\Sigma^{-1}S_n\right)\\ 
  &\qquad + n\mathrm{tr}\left(\Sigma^{-1}(\bar{\mathbf{y}} - \mu)(\bar{\mathbf{y}} - \mu)^T\right) \\
  &= (n-1)\mathrm{tr}\left(\Sigma^{-1}S_n\right)\\ 
  &\qquad + n(\bar{\mathbf{y}} - \mu)^T\Sigma^{-1}(\bar{\mathbf{y}} - \mu). \\
  \end{align*}

## Maximum Likelihood Estimators 
  - Going back to the log-likelihood, we get:
  $$\ell = -\frac{n}{2}\log\lvert\Sigma\rvert - \frac{(n-1)}{2}\mathrm{tr}\left(\Sigma^{-1}S_n\right) - \frac{n}{2}(\bar{\mathbf{y}} - \mu)^T\Sigma^{-1}(\bar{\mathbf{y}} - \mu).$$
  - Since $\Sigma^{-1}$ is positive definite, for $\Sigma$ fixed, the log-likelihood is maximised at
  $$\hat{\mu} = \bar{\mathbf{y}}.$$
  - With extra effort, it can be shown that $-\log\lvert\Sigma\rvert - \frac{(n-1)}{n}\mathrm{tr}\left(\Sigma^{-1}S_n\right)$ is maximised at
  $$\hat{\Sigma} = \frac{(n-1)}{n}S_n = \frac{1}{n}\sum_{i=1}^n(\mathbf{y}_i - \bar{\mathbf{y}})(\mathbf{y}_i - \bar{\mathbf{y}})^T.$$
  - *In other words*: $(\bar{\mathbf{Y}},\hat{\Sigma})$ are the **maximum likelihood estimators** for $(\mu,\Sigma)$.
  
## Maximum Likelihood Estimators

  - Since the multivariate normal density is "well-behaved", we can deduce the usual properties:
    + **Consistency**: $(\bar{\mathbf{Y}},\hat{\Sigma})$ converges in probability to $(\mu,\Sigma)$.
    + **Efficiency**: Asymptotically, the covariance of $(\bar{\mathbf{Y}},\hat{\Sigma})$ achieves the Cramér-Rao lower bound.
    + **Invariance**: For any transformation $(g(\mu),G(\Sigma))$ of $(\mu,\Sigma)$, its MLE is $(g(\bar{\mathbf{Y}}),G(\hat{\Sigma}))$.
    
## Visualizing the likelihood

```{r}
library(mvtnorm)
set.seed(123)

n <- 50; p <- 2

mu <- c(1, 2)
Sigma <- matrix(c(1, 0.5, 0.5, 1), ncol = p)

Y <- rmvnorm(n, mean = mu, sigma = Sigma)
```

## Visualizing the likelihood

```{r}
loglik <- function(mu, sigma, data = Y) {
  # Compute quantities
  y_bar <- colMeans(Y)
  Sn <- cov(Y)
  Sigma_inv <- solve(sigma)
  
  # Compute quadratic form
  quad_form <- drop(t(y_bar - mu) %*% Sigma_inv %*% 
                      (y_bar - mu))
  
  -0.5*n*log(det(sigma)) -
    0.5*(n - 1)*sum(diag(Sigma_inv %*% Sn)) -
    0.5*n*quad_form
}
```

---

```{r}
grid_xy <- expand.grid(seq(0.5, 1.5, 
                           length.out = 32), 
                       seq(1, 3, 
                           length.out = 32))

contours <- purrr::map_df(seq_len(nrow(grid_xy)), 
                          function(i) {
  # Where we will evaluate loglik
  mu_obs <- as.numeric(grid_xy[i,])
  # Evaluate at the pop covariance
  z <- loglik(mu_obs, sigma = Sigma)
  # Output data.frame
  data.frame(x = mu_obs[1],
             y = mu_obs[2],
             z = z)
})
```

## Visualizing the likelihood {.allowframebreaks}

```{r, message = FALSE}
library(tidyverse)
library(ggrepel)
# Create df with pop and sample means
data_means <- data.frame(x = c(mu[1], mean(Y[,1])),
                         y = c(mu[2], mean(Y[,2])),
                         label = c("Pop.", "Sample"))
```


```{r, message = FALSE}
contours %>% 
  ggplot(aes(x, y)) + 
  geom_contour(aes(z = z)) + 
  geom_point(data = data_means) +
  geom_label_repel(data = data_means,
                   aes(label = label))
```

```{r}
library(scatterplot3d)
with(contours, scatterplot3d(x, y, z))
```

```{r, echo = FALSE, eval = FALSE}
x <- seq(0.5, 1.5, length.out = 32)
y <- seq(1, 3, length.out = 32)
z <- matrix(NA, ncol = 32, nrow = 32)
for (i in seq_len(32)) {
  for (j in seq_len(32))
    z[i,j] <- loglik(c(x[i], y[j]), sigma = Sigma)
}
persp(x, y, z, theta = 30, phi = 30, expand = 0.5, col = "lightblue")
```

## Sampling Distributions

  - Recall the univariate case:
    + $\bar{X} \sim N\left(\mu, \sigma^2/n\right)$;
    + $\frac{(n-1)s^2}{\sigma^2} \sim \chi^2(n-1)$;
    + $\bar{X}$ and $s^2$ are independent.
  - In the multivariate case, we have similar results:
    + $\bar{\mathbf{Y}}\sim N_p\left(\mu, \frac{1}{n}\Sigma\right)$;
    + $(n-1)S_n = n\hat{\Sigma}$ follows a *Wishart* distribution with $n-1$ degrees of freedom;
    + $\bar{\mathbf{Y}}$ and $S_n$ are independent.
    
## Wishart Distribution

  - Suppose $\mathbf{Z}_1, \ldots, \mathbf{Z}_n\sim N_p(0, \Sigma)$ are independently distributed. Then we say that 
  $$ W = \sum_{i=1}^n \mathbf{Z}_i\mathbf{Z}_i^T$$
  follows a *Wishart distribution* $W_n(\Sigma)$ with $n$ degrees of freedom.
  - Note that since $E(\mathbf{Z}_i\mathbf{Z}_i^T) = \Sigma$, we have $E(W) = n\Sigma$.
  - From the previous slide: $\sum_{i=1}^n(\mathbf{Y}_i - \mathbf{\bar{Y}})(\mathbf{Y}_i - \mathbf{\bar{Y}})^T$ has the same distribution as $\sum_{i=1}^{{\color{blue}n-1}} \mathbf{Z}_i\mathbf{Z}_i^T$ for some choice of $\mathbf{Z}_1, \ldots, \mathbf{Z}_{n-1}\sim N_p(0,\Sigma)$.
    
## Useful Properties

  - If $W_1 \sim W_{n_1}(\Sigma)$ and $W_2 \sim W_{n_2}(\Sigma)$ are independent, then 
  $$ W_1 + W_2 \sim W_{n_1 + n_2}(\Sigma).$$
  - If $W \sim W_{n}(\Sigma)$ and $C$ is $q\times p$, then
  $$ CWC^T \sim W_n(C\Sigma C^T).$$
  
## Density function

  - Let $\Sigma$ be a fixed $p\times p$ positive definite matrix. The density of the Wishart distribution with $n$ degrees of freedom, with $n \geq p$, is given by
  $$ w_n(A; \Sigma) = \frac{\lvert A\rvert^{(n-p-1)/2}\exp\left(-\frac{1}{2}\mathrm{tr}(\Sigma^{-1}A)\right)}{2^{np/2}\pi^{p(p-1)/4}\lvert\Sigma\rvert^{n/2}\prod_{i=1}^p\left(\frac{1}{2}(n-i+1)\right)},$$
  where $A$ is ranging over all $p\times p$ positive definite matrices.
  
## Eigenvalue density function

  - For a random matrix $A \sim W_{n}(I_p)$ with $n\geq p$, the joint distribution of its eigenvalues $\lambda_1 \geq \cdots \geq \lambda_p$ has density
  $$ C_{n,p}\exp\left(-\frac{1}{2}\sum_{i=1}^p\right)\prod_{i=1}^p \lambda_i^{(n-p-1)/2}\prod_{i < j} \lvert\lambda_i - \lambda_j\vert,$$
  for some constant $C_{n,p}$.
  - We will study this distribution in STAT 7200--Multivariate Analysis I
  