---
title: "Multivariate Normal Distribution"
draft: false
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: STAT 4690--Applied Multivariate Analysis
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Building the multivariate density {.allowframebreaks}

  - Let $Z \sim N(0, 1)$ be a standard (univariate) normal random variable. Recall that its density is given by
  $$ \phi(z) = \frac{1}{\sqrt{2\pi}}\exp\left(-\frac{1}{2}z^2\right).$$
  - Now if we take $Z_1, \ldots, Z_p\sim N(0, 1)$ independently distributed, their joint density is
  \begin{align*}
  \phi(z_1, \ldots, z_p) &= \prod_{i=1}^p \frac{1}{\sqrt{2\pi}}\exp\left(-\frac{1}{2}z_i^2\right)\\
  &= \frac{1}{(\sqrt{2\pi})^p}\exp\left(-\frac{1}{2}\sum_{i=1}^p z_i^2\right)\\
  &=\frac{1}{(\sqrt{2\pi})^p}\exp\left(-\frac{1}{2}\mathbf{z}^T\mathbf{z}\right),\\
  \end{align*}
  where $\mathbf{z} = (z_1, \ldots, z_p)$.
  - More generally, let $\mu\in\mathbb{R}^p$ and let $\Sigma$ be a $p\times p$ positive definite matrix.
    + Let $\Sigma=LL^T$ be the Cholesky decomposition for $\Sigma$. 
  - Let $\mathbf{Z} = (Z_1, \ldots, Z_p)$ be a standard (multivariate) normal random vector, and define $\mathbf{Y} = L\mathbf{Z} + \mu$. We know from last lecture that
    + $E(\mathbf{Y}) = LE(\mathbf{Z}) + \mu = \mu$;
    + $\mathrm{Cov}(\mathbf{Y}) = L\mathrm{Cov}(\mathbf{Z})L^T = \Sigma$.
  - To get the density, we need to compute the inverse transformation:
  $$\mathbf{Z} = L^{-1}(\mathbf{Y} - \mu).$$
  - The Jacobian matrix $J$ for this transformation is simply $L^{-1}$, and therefore
  \begin{align*} 
  \lvert\det(J)\rvert &= \lvert\det(L^{-1})\rvert\\
   &= \det(L)^{-1}\qquad(L\mbox{ is p.d.})\\
   &= \sqrt{\det(\Sigma)}^{-1}\\
   &= \det(\Sigma)^{-1/2}.
   \end{align*}
  - Plugging this into the formula for the density of a transformation, we get
  \begin{align*}
  & f(y_1, \ldots, y_p) = \frac{1}{\det(\Sigma)^{1/2}}\phi(L^{-1}(\mathbf{y} - \mu))\\
  &= \frac{1}{\det(\Sigma)^{1/2}}\left(\frac{1}{(\sqrt{2\pi})^p}\exp\left(-\frac{1}{2}(L^{-1}(\mathbf{y} - \mu))^TL^{-1}(\mathbf{y} - \mu)\right)\right)\\
  &= \frac{1}{\det(\Sigma)^{1/2}(\sqrt{2\pi})^p}\exp\left(-\frac{1}{2}(\mathbf{y} - \mu)^T(LL^T)^{-1}(\mathbf{y} - \mu)\right)\\
  &= \frac{1}{\sqrt{(2\pi)^p\lvert\Sigma\rvert}}\exp\left(-\frac{1}{2}(\mathbf{y} - \mu)^T\Sigma^{-1}(\mathbf{y} - \mu)\right).
  \end{align*}
  
## Example {.allowframebreaks}  

```{r, message = FALSE}
set.seed(123)

n <- 1000; p <- 2
Z <- matrix(rnorm(n*p), ncol = p)

mu <- c(1, 2)
Sigma <- matrix(c(1, 0.5, 0.5, 1), ncol = 2)
L <- t(chol(Sigma))
```


```{r, message = FALSE}
Y <- L %*% t(Z) + mu
Y <- t(Y)

colMeans(Y)
cov(Y)

library(tidyverse)
Y %>% 
  data.frame() %>% 
  ggplot(aes(X1, X2)) +
  geom_density_2d()
```

```{r}
library(mvtnorm)

Y <- rmvnorm(n, mean = mu, sigma = Sigma)

colMeans(Y)
cov(Y)

Y %>% 
  data.frame() %>% 
  ggplot(aes(X1, X2)) +
  geom_density_2d()
```

## Other characterizations

There are at least two other ways to define the multivariate random distribution:

  1. A $p$-dimensional random vector $\mathbf{Y}$ is said to have a multivariate normal distribution if and only if every linear combination of $\mathbf{Y}$ has a *univariate* normal distribution.
  2. A $p$-dimensional random vector $\mathbf{Y}$ is said to have a multivariate normal distribution if and only if its distribution maximises entropy over the class of random vectors with fixed mean $\mu$ and fixed covariance matrix $\Sigma$ and support over $\mathbb{R}^p$. 
  
## Useful properties {.allowframebreaks}

  - If $\mathbf{Y}\sim N_p(\mu, \Sigma)$, $A$ is a $q \times p$ matrix, and $b\in\mathbb{R}^q$, then
  $$A\mathbf{Y} + b \sim N_q (A\mu + b, A\Sigma A^T).$$

  - If $\mathbf{Y}\sim N_p(\mu, \Sigma)$ then all subsets of $\mathbf{Y}$ are normally distributed; that is, write 
    + $\mathbf{Y} = \begin{pmatrix}\mathbf{Y}_1\\\mathbf{Y}_2\end{pmatrix}$, $\mu = \begin{pmatrix}\mu_1\\\mu_2\end{pmatrix}$;
    + $\Sigma = \begin{pmatrix} \Sigma_{11} & \Sigma_{12}\\\Sigma_{21} & \Sigma_{22}\end{pmatrix}$.
    + Then $\mathbf{Y}_1\sim N_r(\mu_1,\Sigma_{11})$ and $\mathbf{Y}_2\sim N_{p-r}(\mu_2,\Sigma_{22})$.
    
  - Assume the same partition as above. Then the following are equivalent:
    + $\mathbf{Y}_1$ and $\mathbf{Y}_2$ are independent;
    + $\Sigma_{12} = 0$;
    + $\mathrm{Cov}(\mathbf{Y}_1, \mathbf{Y}_2) = 0$.
    
## Exercise (J&W 4.3)

Let $(Y_1, Y_2, Y_3) \sim N_3(\mu, \Sigma)$ with $\mu = (−3, 1, 4)$ and
$$ \Sigma = \begin{pmatrix} 1 & -2 & 0\\ -2 & 5 & 0 \\ 0 & 0 & 2\end{pmatrix}.$$
Which of the following random variables are independent? Explain.

  1. $Y_1$ and $Y_2$.
  2. $Y_2$ and $Y_3$.
  3. $(Y_1, Y_2)$ and $Y_3$.
  4. $0.5(Y_1 + Y_2)$ and $Y_3$.
  5. $Y_2$ and $Y_2-\frac{5}{2}Y_1 -Y_3$.
  
## Conditional Normal Distributions i 

  - **Theorem**: Let $\mathbf{Y}\sim N_p(\mu, \Sigma)$, where 
    + $\mathbf{Y} = \begin{pmatrix}\mathbf{Y}_1\\\mathbf{Y}_2\end{pmatrix}$, $\mu = \begin{pmatrix}\mu_1\\\mu_2\end{pmatrix}$;
    + $\Sigma = \begin{pmatrix} \Sigma_{11} & \Sigma_{12}\\\Sigma_{21} & \Sigma_{22}\end{pmatrix}$. 
  - Then the *conditional distribution* of $\mathbf{Y}_1$ given $\mathbf{Y}_2 = y_2$ is multivariate normal $N_r(\mu_{1\mid 2}, \Sigma_{1\mid 2})$, where
    + $\mu_{1\mid 2} = \mu_1 + \Sigma_{12} \Sigma_{22}^{-1}(y_2 - \mu_2)$
    + $\Sigma_{1\mid 2} = \Sigma_{11} + \Sigma_{12} \Sigma_{22}^{-1}\Sigma_{21}$.
    
## Conditional Normal Distributions ii

  - **Corrolary**: Let $\mathbf{Y}_2\sim N_{p-r}(\mu_2, \Sigma_{22})$ and assume that $\mathbf{Y}_1$ given $\mathbf{Y}_2 = y_2$ is multivariate normal $N_r(Ay_2 + b, \Omega)$, where $\Omega$ does not depend on $y_2$. Then $\mathbf{Y} = \begin{pmatrix}\mathbf{Y}_1\\\mathbf{Y}_2\end{pmatrix}\sim N_p(\mu, \Sigma)$, where
    + $\mu = \begin{pmatrix}A\mu_2+b\\\mu_2\end{pmatrix}$;
    + $\Sigma = \begin{pmatrix} \Omega+A\Sigma_{22}A^T & A\Sigma_{22}\\\Sigma_{22}A^T & \Sigma_{22}\end{pmatrix}$. 
    
## Exercise

  - Let $\mathbf{Y}_2\sim N_1(0, 1)$ and assume
  $$\mathbf{Y}_1 \mid \mathbf{Y}_2 = y_2 \sim N_2\left(\begin{pmatrix}y_2+1\\2y_2\end{pmatrix}, I_2\right).$$
  Find the joint distribution of $(\mathbf{Y}_1,\mathbf{Y}_2).$
  
## Another important result {.allowframebreaks}

  - Let $\mathbf{Y}\sim N_p(\mu, \Sigma)$, and let $\Sigma = LL^T$ be the Cholesky decomposition of $\Sigma$.
  - We know that $\mathbf{Z} = L^{-1}(\mathbf{Y} - \mu)$ is normally distributed, with mean 0 and covariance matrix
  $$ \mathrm{Cov}(\mathbf{Z}) = L^{-1}\Sigma(L^{-1})^T = I_p.$$
  - Therefore $(\mathbf{Y} - \mu)^T\Sigma^{-1}(\mathbf{Y} - \mu)$ is the sum of *squared* standard normal random variables.
    + In other words, $(\mathbf{Y} - \mu)^T\Sigma^{-1}(\mathbf{Y} - \mu) \sim \chi^2(p)$.
    + This can be seen as a generalization of the univariate result $\left(\frac{X - \mu}{\sigma}\right)^2\sim\chi^2(1)$.
  - From this, we get a result about the probability that a multivariate normal falls within an *ellipse*:
  $$P\left((\mathbf{Y} - \mu)^T\Sigma^{-1}(\mathbf{Y} - \mu) \leq \chi^2(\alpha;p) \right) = 1 - \alpha.$$
    + We can use this to construct a confidence region around the sample mean.
    